---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
setwd('C:\\ds\\R\\reproducible_research\\proj1')
library(data.table)
act_data <- read.csv('activity.csv', header=TRUE, na.strings='NA')
act_data <- data.table(act_data)
head(act_data)
```

## What is mean total number of steps taken per day?
Filtering out the rows containing valid information and the summing up the total number of steps for each day, we get:
```{r}
act_valid <- act_data[!is.na(act_data$steps), ]
day_act <- act_valid[, list(daysteps=sum(steps)), by=date]
hist(day_act$daysteps, xlab='steps per day', main='Distribution of steps taken per day')
mean_day_steps <- mean(day_act$daysteps)
median_day_steps <- median(day_act$daysteps)
```

The mean and median total number of steps taken per day (excluding days without valid data) is:  
mean = `r as.integer(mean_day_steps)` steps,  
median = `r median_day_steps` steps.

## What is the average daily activity pattern?
From the valid data, calculate the average of steps taken in each interval:
```{r}
int_data <- act_valid[, list(avg_steps=mean(steps)), by=interval]
plot(int_data, type='l', xlab='Interval', ylab='Average steps taken')
interval_with_max_steps <- int_data$interval[which.max(int_data$avg_steps)]
```
The 5-minute interval that on average containing the maximum number of steps is `r interval_with_max_steps`. This is in the morning.

## Imputing missing values
```{r}
numNA <- sum(!complete.cases(act_data))
```
The number of rows with missing values are `r numNA`.

We fill in the missing values with the average values from the same interval averaged across all days:
```{r fillInMissingValues}
act_data_imput <- act_data
bNASteps <- is.na(act_data$steps)
act_data_imput$steps[bNASteps] = int_data$avg_steps[as.integer(as.factor(act_data$interval[bNASteps]))]
```
Using the new dataset with the NA values filled in, we plot the average number of steps taken each day:
```{r}
day_act_imput <- act_data_imput[, list(daysteps=sum(steps)), by=date]
hist(day_act_imput$daysteps, xlab='steps per day', main='Distribution of steps taken per day')
mean_day_steps_imput <- mean(day_act_imput$daysteps)
median_day_steps_imput <- median(day_act_imput$daysteps)
```
The mean and median steps taken per day after filling in the missing values are:  
mean = `r as.integer(mean_day_steps_imput)` steps,   
median = `r as.integer(median_day_steps_imput)` steps.

As we can see, the mean did not change as a result of imputing. This is expected since we filled in the days with missing values using the mean of the valid days. The median did get shifted by a tiny bit. However, the effect of filling in the missing values with mean of valid days is minimal and can be ignored.



## Are there differences in activity patterns between weekdays and weekends?
We use the **dplyr** package to generate a column that identifies each day as a weekday or a weekend day:
```{r}
library(dplyr)
act_data_imput <- tbl_df(act_data_imput)
act_data_imput$weekday <- wday(as.Date(act_data_imput$date, '%Y-%m-%d'))
act_data_imput <- mutate(act_data_imput, wday=(weekday>=2 & weekday<=6))
act_data_imput$wdaytype <- factor(act_data_imput$wday, labels=c('weekend', 'weekday'))
```
Now we separately average the steps for each interval for both week days and the weekend days:
```{r}
wdaysteps <- with(act_data_imput, tapply(steps, list('Interval'=interval, 'DayType'=wdaytype), mean))
df_matrix <- as.data.frame(wdaysteps)
df_matrix$interval <- rownames(df_matrix)
```
Now *df_matrix* contains the average steps taken for each interval on weekdays and weekend days. This data is in a 'messy' state and need to be cleaned up. We use the **tidyr** package to do data cleaning:
```{r}
library(tidyr)
df_matrix <- tbl_df(df_matrix)
df_matrix <- gather(df_matrix, daytype, steps, weekend:weekday) # gather weekday and weekend into one column called daytype
df_matrix <- as.data.frame(df_matrix)
df_matrix$daytype <- as.factor(df_matrix$daytype)
```
Before we plot the steps vs interval, we convert the interval values into time.
```{r}
df_matrix$interval <- as.numeric(df_matrix$interval)
sInterval <- sprintf('%04d', df_matrix$interval)
sInterval <- sub("(\\d+)(\\d{2})", "\\1:\\2:00", sInterval)
head(sInterval)
```
Now that we have the time intervals in correct string format, we convert them into time in minutes:
```{r}
library(lubridate)
df_matrix$tInterval <- as.numeric(as.duration(hms(sInterval) - hms('00:00:00')))/60
head(df_matrix)
```

Now we can plot the average steps taken in each interval for weekdays and weekend days using the lattice package:
```{r}
library(lattice)
xyplot(steps ~ tInterval | daytype, data=df_matrix, layout=c(1, 2), type='l', xlab='Minutes', ylab='Steps')
```
